Constraint Grammar as a SAT problem

### 2015-06-01 paper/nodalida2015.tex
stuff

Constraint Grammar (CG) \cite{karlsson1995constraint} is a relatively
young formalism, born out of practical need for a robust and
language-independent method for part-of-speech tagging.
%It belongs to the reductionist family of parsing
%strategies, along with e.g. Finite-State Intersection Grammar (FSIG) \cite{koskenniemi90}.
In this work, we present CG %\footnote{Roughly compatible with CG-2}
as a Boolean satisfiability (SAT) problem, and describe an implementation
using a SAT solver. This is attractive from several reasons: formal logic is
well-studied, and serves as an abstract language to reason about the
properties of CG. Constraint rules encoded in logic capture richer
dependencies between the tags than standard CG.

@@

As hinted by the previous property, SAT solver doesn't need a fixed
order of the rules.
Applying a rule to a sentence produces a number of clauses,
and those clauses are fed into the SAT solver.
However, in the unordered scheme, some information is lost: the
following rule sets would be treated identically, whereas in the
traditional CG, only the second would be considered as a bad order.

\begin{itemize}
\item [\texttt{1)}] \texttt{REMOVE v IF (-1 det) ;} \\
         \texttt{SELECT v ;}
\item [\texttt{2)}] \texttt{SELECT v ;} \\
         \texttt{REMOVE v IF (-1 det) ;} 
\end{itemize}

@@

We can also emulate order with SAT-CG. To do that, we enter clauses
produced by each rule one by one, and assume the solver state reached
so far is correct. If a conflict is introduced by new clauses, discard
them and move on to the next rule.
By testing against gold standard, we see that this scheme works better
with ready-made CGs, which are written with ordering in mind.
It also runs slightly faster than the unordered version.
%the performance, because the maximisation problems are smaller. 
%We have also experimented with sections, and 
\\

@@

For future work, we intend to apply the SAT solver for corpus-free
conflict testing. Let us illustrate the idea with the same two rules,
\texttt{REMOVE v IF (-1 det)} and \texttt{SELECT v}
in both orders.
Assume we have tag set \texttt{\{det, n, v, sg, pl\}}. Then we
can ask the SAT solver a question if there exists an input which
satisfies our constraints: both rules, applied in the order, remove
something from the input.
SAT solver tells us that with sentence length 2 and maximum 4 readings
for the whole sentence, there are several inputs that
satisfy the constraints for the first rule set, but none that
satisfies the second rule set: for instance

@@

SAT-solvers are nowadays powerful enough to be used for dealing with
Constraint Grammar. A logic-based approach to CG has possible
advantages over more traditional approaches; a SAT-solver may
disambiguate more words, and may do so more precisely. Also, the
SAT-solver requires less rules, and these rules are simpler. 
\todo{proper conclusions for the new stuff}
% The ordering of rules is something we found was incompatible with a logic-based approach. We compensate this by maximising rule applications. Whether this is an advantage or disadvantage remains to be seen. Our initial experimental results are promising.

>>>>

Constraint Grammar (CG) \cite{karlsson1995constraint} is a relatively
young formalism, born out of practical need for a robust and
language-independent method for part-of-speech tagging.
%It belongs to the reductionist family of parsing
%strategies, along with e.g. Finite-State Intersection Grammar (FSIG) \cite{koskenniemi90}.
In this work, we present CG %\footnote{Roughly compatible with CG-2}
as a Boolean satisfiability (SAT) problem, and describe an implementation
using a SAT solver. This is attractive for several reasons: formal logic is
well-studied, and serves as an abstract language to reason about the
properties of CG. Constraint rules encoded in logic capture richer
dependencies between the tags than standard CG.

@@

As hinted by the previous property, SAT solver doesn't need a fixed
order of the rules.
Applying a rule to a sentence produces a number of clauses,
and those clauses are fed into the SAT solver.
However, in the unordered scheme, some information is lost: the
following rule sets would be treated identically, whereas in the
traditional CG, only the first would be considered as a bad order.

\begin{itemize}
\item [\texttt{1)}] \texttt{SELECT v ;} \\
         \texttt{REMOVE v IF (-1 det) ;} 
\item [\texttt{2)}] \texttt{REMOVE v IF (-1 det) ;} \\
         \texttt{SELECT v ;}
\end{itemize}

@@

We can also emulate order with SAT-CG. To do that, we enter clauses
produced by each rule one by one, and assume the solver state reached
so far is correct. If a new clause introduces a conflict with
previous clauses, we discard it and move on to the next rule.
By testing against gold standard, we see that this scheme works better
with ready-made CGs, which are written with ordering in mind.
It also runs slightly faster than the unordered version.
%the performance, because the maximisation problems are smaller. 
%We have also experimented with sections, and 
\\

@@

As future work, we intend to test for conflicts
without using a corpus.
Let us illustrate the idea with the same two rules,
\texttt{SELECT v} and \texttt{REMOVE v IF (-1 det)}
in both orders.
Assume we have the tag set \texttt{\{det, n, v, sg, pl\}}, 
and we want to find if there exists an input such that
 both rules, applied in the given order, remove something from the input.
% This is another constraint solving problem, hence we can model it in SAT.
There are no inputs that satisfy the requirement with first order,
but several that work with the second, for instance:

@@

SAT-solvers are nowadays powerful enough to be used for dealing with
Constraint Grammar. A logic-based approach to CG has possible
advantages over more traditional approaches; a SAT-solver may
disambiguate more words, and may do so more precisely, capturing
more dependencies between tags.
We experimented with both ordered and unordered rules, and found the ordered
scheme to work better with previously written grammars.

% Also, the SAT-solver requires less rules, and these rules are simpler. 
% The ordering of rules is something we found was incompatible with a logic-based approach. We compensate this by maximising rule applications. Whether this is an advantage or disadvantage remains to be seen. Our initial experimental results are promising.
